---
title: "Computational Musicology"
author: "Dewi Timman"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: "spacelab"
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(spotifyr)
library(plotly)
library(compmus)
```

```{r corpus setup, include=FALSE}
corpus <- get_playlist_audio_features("", "3jblLN3iqoULSfQGHAFYVe")
sleeping_at_last_tracks <- get_artist_audio_features("sleeping at last")
corpus$sleeping <- ifelse(corpus$track.id %in% sleeping_at_last_tracks$track_id, 1, 0)
```


### Corpus

[My corpus](https://open.spotify.com/playlist/3jblLN3iqoULSfQGHAFYVe?si=262d717ca4354192) is a playlist with around 3500 songs.
What do all the songs have in common?
They are all songs I like.
It is a very diverse list with songs ranging from film music to _Shawn Mendes_ to _The Sidh_ to _Maria Mena_.
But there is one specific artist I listen to significantly more than any other artist: _Sleeping At Last_.
_Sleeping At Last_ is an artist that has all kinds of projects for himself.
For example, he makes songs about upcoming astronomical events (_Astronomy_ series), or about involuntary human development (_Atlas II_ series).
He also has a podcast on Spotify about how he makes his songs.
_Sleeping At Last_ has a thin line with film music.
His music can be heard often in films and series.
But his music also comes in contact with more popular artists: he has made some covers.
Because of these things, he is easy to compare to other songs in the playlist.
However, I am not interested in how _Sleeping At Last_ compares to the other artists, but I am interested in how he differs from the other songs in the playlist.
What makes _Sleeping At Last_ _Sleeping At Last_?
Why is this the artist I listen to most often?
Are there other artists with the same characteristics?

I think that to get an even deeper understanding of the style of _Sleeping At Last_, there are a few songs that are most interesting to look at.
First, one or more _astronomy_ songs from _Sleeping At Last_.
These are the songs that are very typical for _Sleeping At Last_.
Both in the sense of the subject and the kind of music and instruments.
Moreover, it would be nice to look at one or more of the songs that are included in a film or TV series and compare it to another song in that film or TV series to see what is so special about this song being there.
Furthermore, one or more of his covers can be compared to the original to see how the styles differ and get to know more about this specific style of _Sleeping At Last_.

***

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/3jblLN3iqoULSfQGHAFYVe?utm_source=generator" width="100%" height="352" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" loading="lazy"></iframe>

### Valence, acousticness, danceability and instrumentalness

```{r track-level features plot, echo=FALSE}
plot <- ggplot(corpus) +
  geom_point(aes(x=valence, y=acousticness, size=danceability, color=instrumentalness), alpha=0.4, shape=21) +
  geom_rug(aes(x=valence, y=acousticness, color=instrumentalness)) +
  facet_grid(cols = vars(sleeping), labeller=as_labeller(c(`0`='Other songs', `1`='Sleeping At Last songs'))) +
  labs(title='Different track-level features of Sleeping At Last songs\ncompared to other songs in the chosen playlist') +
  scale_color_distiller(palette='PuOr') + 
  scale_x_continuous(breaks=c(0,0.5,1),limits=c(0,1)) +
  scale_y_continuous(breaks=c(0,0.5,1), limits=c(0,1))

theme <- theme_minimal() +
  theme(panel.grid.minor=element_blank(),
        axis.line=element_line(), 
        plot.title = element_text(hjust = 0.5))

#ggplotly(plot + theme)
plot + theme
```

***

In this graphic, the track-level features valence, acousticness, instrumentalness and danceability are shown for the _Sleeping At Last_ songs in the playlist compared to the other songs in the playlist.
The graphic shows that _Sleeping At Last songs_ have a higer acousticness than the other songs in general, whereas the valence and danceability tend to be lower.
In addition, the proportion of instrumental songs seems to be bigger in the _Sleeping At Last_ songs than in the other songs.


### Tempo of the song _Dark Horse_ by _Katy Perry_ and _Sleeping At Last_

```{r chroma corpus setup, echo=FALSE}
dark_horse_katy <- get_tidy_audio_analysis("4jbmgIyjGoXjY01XxatOx6") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)
dark_horse_sleeping <- get_tidy_audio_analysis("1i5PW20LSYwCQMjVQgSXVM") %>%
  select(segments) %>%
  unnest(segments) %>%
  select(start, duration, pitches)
```

```{r chroma plot, echo=FALSE}
dark_horse_dist <- compmus_long_distance(
  dark_horse_katy %>% mutate(pitches = map(pitches, compmus_normalise, "manhattan")),
  dark_horse_sleeping %>% mutate(pitches = map(pitches, compmus_normalise, "manhattan")),
  feature = pitches,
  method = "aitchison"
)

dark_horse <- dark_horse_dist %>%
  mutate(dark_horse_sleeping = xstart + xduration / 2,
         dark_horse_katy = ystart + yduration / 2) %>%
  ggplot(aes(x=dark_horse_sleeping, y=dark_horse_katy, fill=d)) +
    geom_tile(aes(width=xduration, height=yduration)) +
    coord_fixed() + 
    scale_x_continuous(breaks=c(16,51,89,123,163,192,209,221), labels=c("I knew you were", "So you wanna play with magic?", "Mark my words", "So you wanna play with magic?", "So you wanna play with magic?", "There's no going back (first time)", "There's no going back (last time)", "")) + 
    scale_y_continuous(breaks = c(15, 45, 80, 110, 139, 183, 210, 216), labels=c("I knew you were", "So you wanna play with magic?", "Mark my words", "So you wanna play with magic?", "Uh, she's a beast", "So you wanna play with magic?", "There's no going back", "")) + 
    scale_fill_viridis_c(option = "E", guide = "none") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(x = "Sleeping At Last", y = "Katy Perry")

dark_horse
```

***

In this plot the tempo differences between _Katy Perry_ and _Sleeping At Last_ are shown for the song _Dark Horse_.
It can be seen that, although there are some points in the song where both artists have the same tempo, _Katy Perry's_ tempo is a bit faster in general.
At the end of the song, _Sleeping At Last_ deviates from the original song by skipping over the bridge (starting with "Uh, she's a beast") and repeating 'There's no going back' multiple times.
