---
title: "Computational Musicology"
author: "Dewi Timman"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: "spacelab"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(spotifyr)
library(plotly)
```

```{r corpus setup, include=FALSE}
corpus <- get_playlist_audio_features("", "3jblLN3iqoULSfQGHAFYVe")
sleeping_at_last_tracks <- get_artist_audio_features("sleeping at last")
corpus$sleeping <- ifelse(corpus$track.id %in% sleeping_at_last_tracks$track_id, 1, 0)
```


Introduction
=====================================================

Corpus
---------------------------------------------------------------
[My corpus](https://open.spotify.com/playlist/3jblLN3iqoULSfQGHAFYVe?si=262d717ca4354192) is a playlist with around 3500 songs. 
What do all the songs have in common? They are all songs I like. 
It is a very diverse list with songs ranging from film music to *Shawn Mendes* to *The Sidh* to *Maria Mena*. But there is one specific artist I listen to significantly more than any other artist: *Sleeping At Last*. 
*Sleeping At Last* is an artist that has all kinds of projects for himself. For example, he makes songs about upcoming astronomical events (*Atronomy* series), or about involuntary human development (*Atlas II* series). He also has a podcast on Spotify about how he makes his songs. 
*Sleeping At Last* has a thin line with film music. His music can be heard often in films and series. 
But his music also comes in contact with more popular artists: he has made some covers. Because of these things, he is easy to compare to other songs in the playlist. 
However, I am not interested in how *Sleeping At Last* compares to the other artists, but I am interested in how he differs from the other songs in the playlist. 
What makes *Sleeping At Last* *Sleeping At Last*? Why is this the artist I listen to most often? Are there other artists with the same characteristics?

I think that to get an even deeper understanding of the style of *Sleepnig At Last*, there are a few songs that are most interesting to look at. 
First, one or more *astronomy* songs from *Sleeping At Last*. These are the songs that are very typical for *Sleeping At Last*. Both in the sense of the subject and the kind of music and instruments. 
Moreover, it would be nice to look at one or more of the songs that are included in a film or TV series and compare it to another song in that film or TV series to see what is so special about this song being there. 
Furthermore, one or more of his covers can be compared to the original to see how the styles differ and get to know more about this specific style of *Sleeping At Last*. 

Track-level features {data-navmenu=Visualisation}
=====================================================
### Track-level features

```{r track-level features plot, echo=FALSE}
plot <- ggplot(corpus) +
  geom_point(aes(x=valence, y=acousticness, size=danceability, color=instrumentalness), alpha=0.4, shape=21) +
  geom_rug(aes(x=valence, y=acousticness, color=instrumentalness)) +
  facet_grid(cols = vars(sleeping), labeller=as_labeller(c(`0`='Other songs', `1`='Sleeping At Last songs'))) +
  labs(title='Different track-level features of Sleeping At Last songs\ncompared to other songs in the chosen playlist') +
  scale_color_distiller(palette='PuOr') + 
  scale_x_continuous(breaks=c(0,0.5,1),limits=c(0,1)) +
  scale_y_continuous(breaks=c(0,0.5,1), limits=c(0,1))

theme <- theme_minimal() +
  theme(panel.grid.minor=element_blank(),
        axis.line=element_line(), 
        plot.title = element_text(hjust = 0.5))

plot1 <- ggplotly(plot + theme)
plot1
```


> In this graphic, the track-level features valence, acousticness, instrumentalness and danceability are shown for the *Sleeping At Last* songs in the playlist compared to the other songs in the playlist. The graphic shows that *Sleeping At Last* songs have a higer acousticness than the other songs in genereal, whereas the valence and danceability tend to be lower. In addition, the proportion of instrumental songs seems to be bigger in the *Sleeping At Last* songs than in the other songs. 

Chroma features {data-navmenu=Visualisation}
=====================================================

### Chart 2

```{r, include=FALSE}

```

